# Effective use of graphs 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment=NA, fig.align='center', fig.width=5.5,fig.height=5.5)
```

## General principles

* Focus the eye on features that are important
* Avoid distracting features
* Lines that are intended to attract attention can be
thickened
* Where points should be the focus, make them large & dark
    + It often makes sense to de-emphasize the axes.
* If points are numerous and there is substantial overlap,
  use open symbols, and/or use symbols that have some
  degree of transparency.
* Different choices of color palettes are effective for different purposes.
* Bear in mind that the eye has difficulty in focusing
simultaneously on widely separated colors that are close
together on the same graph.

## Varying time intervals --- show rates, not counts

A graph that was essentially the solid segmented solid line in 
Figure \@ref(fig:Nobel) appeared in @national1975science 
"Science Indicators, 1974". The segmented line gives
a highly misleading impression for the four years 1971-1974, as
opposed to earlier points, where numbers are totals over decades.
It joins up a final point that is a different measure from earlier
points.  

The gray dots, and the axis on the right, show rates per year,
thus comparing like with like.

```{r cap2, echo=FALSE}
cap2 <- "The black line shows numbers of US Nobel prizes, for given time intervals. The gray dots. with the right axis label, show average per year."
```

```{r Nobel, fig.width=6.5, fig.asp=0.575, fig.show="hold", out.width="80%", fig.pos='H', fig.cap=cap2}
par(mar=c(4.1,3.1,2.1,3.1),mgp=c(2.25,0.5,0))
n2 <- c(1,1,4,12,15,29,27,11)
n1 <- c(1,1,4,12,15,29,27,39)
plot(1:8, n2, xaxt="n", xlab="", ylim=range(n1), lwd=2,
     ylab="Number of Nobel prizes", type="l", fg='gray')
axis(1, at=1:8, labels=paste0(seq(from=1901, to=1971, by=10),"-"), 
     cex.axis=0.9, gap.axis=0.15, col="gray")
axis(1, at=1:8, labels=paste0(c(seq(from=1910, to=1970, by=10),1974)),
     line=0.8, lty=0, cex.axis=0.9, col="gray")
axis(1, at=1:7, labels=FALSE, col="gray")
mtext(side=3, line=0.5, at=0.5, "Total prizes (black line)", adj=0, cex=1.25)
mtext(side=3, line=0.5, at=8.5, "Nobel prizes per year (gray dots)", col='gray60', adj=1, cex=1.25)
points(1:8, n1, lwd=8, col="gray", pch=1)
lines(7:8, tail(n1,2), lty=2,lwd=2, col="gray")
axis(4, at=(0:4)*10, labels=paste(0:4), 
     col.ticks="gray", col='gray')
mtext(side=4, line=1.8, "Nobel prizes per year", col='gray60')
```

The same principle applies for intervals of measures
other than time --- for example of length or volume.

## Banking --- the importance of aspect ratio

```{r cap1, echo=FALSE}
cap1 <- "The same data are used for both graphs.  The pattern that is not
obvious in Panel A is very obvious in Panel B"
```


```{r Banking, fig.width=8, fig.height=3.5, out.width="100%", fig.show="hold", fig.pos="H", fig.cap=cap1}
suppressPackageStartupMessages(library(latticeExtra))
gph1 <- xyplot(sin((1:30)*0.92) ~ I((1:30)*0.92), xlab="", ylab="",
 ##      main=list("A: Pattern is hidden to view", 
##                 x=0.1, y=-0.5, just='left', font=1),
scales=list(y=list(at=c(-1,0,1))))
gph2 <- xyplot(sin((1:30)*0.92) ~ I((1:30)*0.92), aspect='xy', 
               xlab="", ylab="", ylim=c(-9,9),
##               main=list("B: Pattern is now obvious", 
##                 x=0.1, y=-0.5, just='left', font=1)
scales=list(y=list(at=seq(from=-8,to=8,by=2)))
)
update(c("A: Pattern is hidden to view"=gph1,
"B: Pattern is now obvious"=gph2, y.same=FALSE), between=list(x=0.5,y=0))
```

Patterns of change on the horizontal scale that it is
important to identify should bank at an angle of roughly 45^o^
above or below the horizontal.  Angles in
the approximate range 20^o^ to 70^o^ may be satisfactory,
and the aspect ratio should be chosen accordingly.

## Scales that show changes by equal factors

Figure \@ref(fig:animals) shows two plots of the same data.
Panel A plots brain weight (grams) against body weight (kilograms),
for 28 "animals".  Panel B plots the same data, but now equal distances
on each scale show changes by the same factor (i.e., change in relative
weight).  

```{r cap3, echo=FALSE}
cap3 <- "Panel A plots brain weight (grams) against body weight 
(kilograms), for 28 'animals'.  Panel B plots the same data, 
with relative weight scales, i.e., equal distances on each scale 
show changes by the same factor."
```

```{r animals, fig.width=8, fig.height=3.5, out.width="100%", fig.show="hold", fig.pos="H", fig.cap=cap3}
Animals <- MASS::Animals
gph1 <- xyplot(brain ~ body, data=Animals,
        xlab="Body weight (k)", ylab="Brain weight (g)", 
       main=list("A: Scales show weights", x=0.1, y=-0.5, just='left', font=1))
print(gph1, position=c(0,0,0.525,1), more=TRUE)
labs <- c('Asian elephant','African elephant','Human','Dipliodocus',
          'Triceratops','Brachiosaurus','Mouse')
atx <- c(.1,1,10,100,1000,10000,100000)
aty <- c(1,10,100,1000,10000)
Animlabs <- subset(Animals, rownames(Animals)%in%labs)
gph2 <- xyplot(brain ~ body, data=Animals,
scales=list(log='e', x=list(at=atx,labels=paste(atx)),
                   y=list(at=aty,labels=paste(aty))),
       xlab="Body weight (kg)", ylab="  ",
       main=list("B: Relative weight scales", 
                 x=0.1, y=-0.5, just='left', font=1)) +
  latticeExtra::layer(data=Animlabs, 
    panel.text(log(body),log(brain)+c(-0.1,-0.12,0,0.2,0.15,0,0), 
               cex=0.75, labels=rownames(Animlabs), pos=c(c(rep(2,5),4,2))))
print(gph2, position=c(0.475,0,1,1))
```

Often, when measurement data span a large range (e.g., a change
from smallest to largest by a factor of 100 to 1 or more),
it is a relative amount scale that is appropriate.^[Technically,
such scales are known as logarithmic scales.]

## Different graphs serve different purposes {#track}

The line in Figure \@ref(fig:resid)A shows the broad overall pattern, while
Figure \@ref(fig:resid)B shows how that pattern needs to be tweaked to more
closely reflect the data.

```{r cap6, echo=FALSE}
cap6 <- "Panel A plots world record Time (as of 2006, in minutes) against Distance
(in kilometers), for field races.  On both the `x` and `y` axes, the scale
is one on which equal distances on the scale correspond to equal
relative changes.  Panel B plots deviations from the fitted line, otherwise
known as residuals, against Distance. The deviations are approximate relative
differences from the line.  Thus a 0.05 difference is a difference that amounts
to 5\\% of the time predicted by the line."
```

```{r resid, fig.width=8, fig.height=3.25, out.width="100%", fig.show="hold", fig.pos="H", fig.cap=cap6}
atx <- c(.1,.3,1,3,10,30,100)
aty=c(.1,.3,1,3,10,30,100,300)
track <- subset(DAAG::worldRecords, roadORtrack=='track')
gph1 <- xyplot(Time ~ Distance, type=c('p','r'), data=track,
        xlab="Distance (km)", ylab="Time (min)", 
       scales=list(log='e', x=list(at=atx,labels=paste(atx)),
                   y=list(at=aty,labels=paste(aty))),
       main=list("A: Time versus Distance", x=0.1, y=-0.5, just='left', font=1))  +
 latticeExtra::layer(panel.axis(side='top', at=log(c(.2,2,20)), 
 labels=c('200m',"2km",'20km'), text.col="gray50", line.col='gray50',line.lty=2,
 half=FALSE, rot=0))
print(gph1, position=c(0,0,0.5,1), more=TRUE)
track.res <-residuals(lm(log(Time)~log(Distance), data=track))
gph2 <- xyplot(track.res ~ Distance, type=c('p','r'), data=track,
       scales=list(x=list(log='e', at=atx,labels=paste(atx))),
       xlab="Distance (km)", ylab="Deviations from fitted line",
       main=list("B: Residuals versus Distance", x=0.1, y=-0.5, just='left', font=1)) +
 latticeExtra::layer(panel.axis(side='top', at=log(c(.2,2,20)), 
 labels=c('200m',"2km",'20km'), text.col="gray50", line.col='gray50',line.lty=2,
 half=FALSE, rot=0))
print(gph2, position=c(0.5,0,1,1))
```

Notice, in Panel A, the use of scales for which which equal distances on 
the scale correspond to equal relative changes.  This is achieved by
specifying a logarithmic scale.  

In Figure \@ref(fig:resid), the line looks to be a good fit.  The range
of times is however large, from just under 10 seconds to close to 11.5  hours.
All except the largest difference from the fitted line are a less than 7%
change, and are not at all obvious in Panel A.  There is a very clear pattern
of systematic differences in Panel B that reflects differences in human 
physiology, very likely between the athletes who excel at the different
distances.

The line can be interpreting as implying a 13% increase in the time 
per unit distance for every unit increase in the distance.  The units 
may for example be units of 100 meters, or kilometers.  Panel B
indicates that the pattern of increase moves down to a local minimum
at around 200 meters, up to a local maximum at around 1 kilometer, down
again to a local minimum at around 20 kilometers, and then steadily up
again.

### Relative distance scales {-}

Figure \@ref(fig:lnscales) shows different "equal physical distance
along the scale" labels that might be used for the relative
`Distance` scale in Figure \@ref(fig:resid)} in Subsection 
\@ref{track}.  The physical distances along the scale are the 
logarithms of the `Distance` values. 

```{r cap5, echo=FALSE}
cap5 <- "Different labelings, all with tick marks at the same
relative distance apart, are shown for the `Distance` scale.
The multipliers for the `Distance` values that are plotted are,
starting at the bottom, 2, 4, 5, and 10."
```

```{r lnscales, fig.width=4.5, fig.height=3.75, out.width="48%", fig.show="hold", fig.pos="H", fig.cap=cap5, echo=FALSE, fig.align='center'}
atx <- c(.1,.3,1,3,10,30,100)
track <- subset(DAAG::worldRecords, roadORtrack=='track')
rx <- range(track$Distance)
atx10 <- 10^seq(from = floor(log10(rx[1])), to=ceiling(log10(rx[2])))
atx2 <- 2^seq(from = floor(log(rx[1],base=2)), to=ceiling(log(rx[2],base=2)))
atx4 <- 4^seq(from = floor(log(rx[1],base=4)), to=ceiling(log(rx[2],base=4)))
atx5 <- 5^seq(from = floor(log(rx[1],base=5)), to=ceiling(log(rx[2],base=5)))
plot(log(rx), c(0,9), type='n', mgp=c(0,0,0), axes=FALSE, xlab='', ylab='')
axis(1, at=log(atx10), labels=paste(atx10), pos=9)
axis(1, at=log(atx5), labels=paste(atx5), pos=6)
axis(1, at=log(atx4), labels=paste(atx4), pos=3)
axis(1, at=log(atx2), labels=paste(atx2), pos=0)
axis(2, at=c(0,3,6,9), lwd=0, lwd.ticks=0,
     labels=expression('' %*% 2,'' %*% 4,'' %*% 5,'' %*% 10), las=1)
```


## Helpful web links are: 

* Good & bad graphs (Murrell, lecture notes)^[<https://www.stat.auckland.ac.nz/~ihaka/120/Lectures/lecture03.pdf>]
* A short tour of bad graphs (Schwartz)^[<www.stat.sfu.ca/~cschwarz/Stat-650/Notes/PDF/ChapterBadgraphs.pdf>]
* Misleading graphs^[<www.statisticshowto.com/misleading-graphs/>]
    + Color choices^[<www.stonesc.com/pubs/Expert%20Color%20Choices.pdf>]
    + Color Brewer^[<www.statisticshowto.com/misleading-graphs/>]
