# Medicine and health

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment=NA, fig.align='center', fig.width=5.5,fig.height=5.5)
```

## Useful sources of advice and information

Here are noted web resources that, for many issues of major
interest, provide carefully collated assessments that are
based on a critical examination of the whole range of evidence
that the authors could identify.  There may be many 
published studies, not all of equal quality, that bear on a medical 
issue.  Or there may be one, or a small number, of large-scale high 
quality trials that are used as a basis for judgment.  The updating 
of assessments as new evidence emerges is a major challenge.

Resources that will be noted are

* [Harding Center for Risk Literacy](https://www.hardingcenter.de/en)
* The [Cochrane Center](<https://www.cochrane.org/>)
* [Winton Centre for Risk and Evidence  Communication](<https://wintoncentre.maths.cam.ac.uk/>)

[Note also the US Consumer health information site](<https://medlineplus.gov/>)


### The Harding Center for Risk Literacy {-} 

There is extensive informative content on the [Harding Center for Risk Literacy site](https://www.hardingcenter.de/en)^[<https://www.hardingcenter.de/en>] site.  Note in particular

* Medical Fact Boxes.  Look under **Transfer and Impact | Fact Boxes**
* Covid-19. Look under **Transfer and Impact | Corona pandemic**
* Consumer Empowerment. **Transfer and Impact | Consumer Empowerment**
* Risk and Evidence Communication. Look under **Research | Risk and Evidence Communication**
* VisRisk - Visualization and Communication of Complex Evidence in Risk Assessment. Look under **Research | VisRisk**
* Horizon2020 Project FORECEE (on balancing female cancer risk against
the risk arising from false-positive alarms and overdiagnosis).
 Look under **Research | FORECEE Project**
* Drone Risks Project. Look under **Research | Drone Risks Project**
* Publications.  Look under **The Harding Center | Publications**
* In the Media. Look under  **The Harding Center | In the Media**


### Harding Center Medical Fact Boxes {-}

[Medical fact boxes](https://www.hardingcenter.de/en/transfer-and-impact/fact-boxes)^[<https://www.hardingcenter.de/en/transfer-and-impact/fact-boxes>] provide visual and tabular summaries of the current 
"best" evidence, from randomized controlled trials.  The comparison 
may be with a placebo, or with an alternative that is known to be
effective. Detailed references are given.  Where available, reliance 
is on Cochrane studies.  Available fact boxes, as of August 2021, are:

* Vaccines
* General health checks
* Dietary Supplements - Selenium
* Low back pain
* Antibiotics
* Early detection of cancer: breast, colon, prostate
* Cardiovascular Diseases
* Osteoarthritis of the knee
* Tonsil surgery

Scroll down the home page (<https://www.hardingcenter.de/en>) to find the
headings **Risk Literate** (with a link to a risk quiz), and 
**Quick Risk Test** (with a link to a test that is targeted at medical 
students and medical professionals.  On the development of this latter 
test, see @cokely2012measuring.

### The Cochrane Center^[<https://www.cochrane.org/>] {-}

The Cochrane Center's mission is "to promote evidence-informed health decision-making by producing high-quality, relevant, accessible systematic 
reviews and other synthesized research evidence." They rely heavily on
meta-analyses, looking for the balance of evidence across all relevant
studies.

### Winton Centre for Risk and Evidence  Communication^[<https://wintoncentre.maths.cam.ac.uk/>] {-}

## Screening for cancer --- randomized trials

### PSA Screening for Prostate Cancer, & more {-}

Numbers (rounded) in the following table are from a Harding Center
fact box. They are for men 50 years or older who either did or did not 
participate in prostate cancer screening, using the PSA test, for 16 years.^[<https://www.hardingcenter.de/en/early-detection-of-cancer/early-detection-of-prostate-cancer-with-psa-testing>]

|| 1000 men, No screening | 1000 men, Screening
| --- | ------- | ------- 
| Deaths (prostate cancer) | 12 | 10
|||
| Biopsy & false alarm | 0 | 155
| Unnecessary treatment | 0 | 51

About 10 out of every 1,000 men with screening, and 12 out of every 1,000 men without screening, died from prostate cancer within 16 years. This means that 2 out of every 1,000 people could be saved from death from prostate cancer by early detection using PSA testing. Deaths from any cause were around 322 in both groups.

Numbers for benefits are based on four studies with about 77,000 participants (progressive cancer), four studies with about 472,000 participants (overall mortality), and eleven studies with about 619,000 participants (prostate cancer specific mortality). The numbers for harms are based on seven studies with approximately 128,000 participants (false-positive results within three to six participations in PSA testing for early detection) and nine studies with approximately 274,000 participants (over-diagnosis and over-treatment).  See the web site for references to the studies.

Unlike the biopsies that may follow a positive PSA test, the PSA test has
no direct potential to cause physical harm.  Harm results from an
undue readiness to use the test result as a reason for further 
potentially harmful testing and treatment.  "Wait and watch" is often
the preferred strategy.

See also: @levitin_2015, Chapter 6, and 
[How Patients Think, and How They Should](https://www.nytimes.com/2011/10/09/books/review/your-medical-mind-by-jerome-groopman-and-pamela-hartzband-book-review.html)^[<https://www.nytimes.com/2011/10/09/books/review/your-medical-mind-by-jerome-groopman-and-pamela-hartzband-book-review.html>]

### Breast cancer screening --- a contested area {-}

See, e.g., @raichand2017conclusions. The review starts with the comment:
    
> The recent controversy about using mammography to screen for breast cancer 
based on randomized controlled trials over 3 decades in Western countries has 
not only eclipsed the paradigm of evidence-based medicine, but also puts health decision-makers in countries where breast cancer screening is still being considered in a dilemma to adopt or abandon such a well-established screening modality.

The short summary from the 
[Harding Center Fact Box for Mammography Screening](https://www.hardingcenter.de/en/early-detection-breast-cancer-mammography-screening), referring to women 50 years [a few trials looked at
women aged>=40] and older who either did or did not participate in 
mammography screening for approximately 11 years, reads

> Mammography reduced the number of women who died from breast
cancer by 1 out of every 1000 women.  It had no effect on the 
number of women who died from any type of cancer.  Among all
women taking part in screening, some women with non-progressive
cancer were over-diagnosed and received unnecessary treatment.^[<https://www.hardingcenter.de/en/early-detection-breast-cancer-mammography-screening>] 

Those who are convinced of 
the virtues of screening for a wide age group may wish  to challenge 
the evidence as presented at the time of writing.  On what grounds, 
however?  It might be argued that

* Improvements in medical procedures since the time of the trials
on which the Fact Boxes are based have changed the balance of risk.
* The trials do not apply to New Zealand conditions.  

Both cases require justification.

## Randomized Controlled Trials vs other study types {#ss:rct}

Two types of study are widely used in medical and other contexts ---
randomized controlled trials (RCTs), and population-based studies. 
These can, in both cases, be broken down into further sub-types. 
There may be elements of both these types of studies.

### Randomized Controlled Trials --- the gold standard? {-}

* Use a random mechanism to assign to assign to treatment as
    against control --- in a medical screening study to
    "screen" or "not screen"
    + The aim is to ensure that apples are compared with apples
    + Treatment and control must otherwise be treated in the
    same way.
* There must be strict adherence to a protocol
    + Minor departures that may, e.g., allow unconscious bias
    in the way that results from the different groups of participants
    are measured, can invalidate results.
* Results apply, strictly, to those who meet the trial entry criteria
    + This may limit relevance to the general population

Especially in medical trials, think carefully about the outcome measure.

* In a screening trial, e.g., for prostate cancer, there are
risks both for those who test positive, and for
those who test negative.
    + The process used to check for cancer may itself bring
a smaller or larger element of risk.
    + Positives may be false positives, leading to more invasive
    checks which may themselves carry a risk.  Thus, for prostate
    cancer, a positive PSA test is likely to lead to a biopsy that
    itself has been estimated to carry a 5% risk of serious side
    effects, with a much higher proportion of less serious effects
    [@levitin_2015, p.245]
    + Some slow growing cancers may be better left untreated, 
    rather than exposing the patient to a treatment that may itself 
    do serious damage. 
* Thus, think carefully about the choice of outcome measure. It is not
enough to show that a screening program will pick up otherwise 
undiagnosed cancers.

For a helpful animated summary of some of the key issues, see:  
<https://www.youtube.com/watch?v=Wy7qpJeozec>

### A note in passing: HiPPO decisions vs A/B testing {-}

Randomized studies are widely used outside of medicine.
Randomization is a key component of the way that Google and
others test out, e.g., the effect of different web page layouts.

* HiPPO = "Highest paid person in the Office."
* The term "A/B testing" is sometimes used to refer to randomized
testing of alternatives. 

A/B testing helped propel Obama into office!  An experiment was
conducted that involved 15 million people, or about 25%, from
its email list.  The signup forms had one of nine different 
combinations of images with words on which recipients were
invited to click, thus:
\begin{tabular}{lccc}
& Learn more & Join us & Sign up now \\
Obama photo & \ding{56} & \ding{56} & \ding{56} \\
BW photo of Obama family & \ding{52} & \ding{56} & \ding{56} \\
Obama speaking & \ding{56} & \ding{56} & \ding{56} \\
\end{tabular}

The black and white photo of the Obama family, with the words
"Learn more", generated the most clicks.

@young2014improving gives an account of A/B testing as it might 
be used for improving library user experience.
    
### Population studies --- groups must be broadly comparable {-}

* Adjust prunes to look like apples (is it possible?)
* Can one ever be sure that the adjustments do the job?
* Potential for biases is greater than for RCTs

Where a treatment is compared with a control group, the
idea is to use a regression type approach to adjust for
differences in such variables or factors as age, sex,
socioeconomic status, and co-morbidities.
"Propensity score" approaches try to summarize such group
differences in a single variable (or, in principle, two
or more variables) that measure the propensity
to belong to the treatment as opposed to the control group.
While their effectiveness for this purpose may be doubted,
they can be used to provide insightful graphs that check
the extent to which the groups are broadly comparable on
the variables and/or factors used to adjust for differences.

### Issues for all types of study {-} 

What are the relevant outcome measures?

* e.g., cancer -- malignancies found & removed, or deaths
    + deaths from cancer, or from all causes (for some individuals,
    the treatment may be more damaging in it medium to long term effect
    than the cancer)

Care is required to deal with survivor, as well as other, biases. 

## Avoid, or expose infants to peanuts?

Clinical practice guidelines introduced  in or around the
year 2000 had "recommended the exclusion of 
allergenic foods from the diets of infants at high risk for 
allergy, and from the diets of their mothers during pregnancy 
and lactation."  

It was then a surprise to find that
the prevalence of peanut allergy has substantially increased
in the recent past, doubling in Europe between 2005 and 2015,
suggesting that advice given to parents of young children to 
avoid foods containing peanuts may have been counterproductive.
This reassessment was supported, at least for infants who
at four months had either severe eczema or food allergy or 
both, and thus were at high risk of developing a peanut 
allergy, by the LEAPS study reported in @du2015randomized.

As noted, the LEAPS study was limited to infants who at four 
months had either severe eczema or food allergy or both.  
Infants were stratified into two groups following a skin-prick 
test, with each group then randomized between those exposed 
to peanut extract, and those not exposed.

Among 530 infants in the  population who initially had negative results on the skin-prick test, the prevalence of peanut allergy at 60 months of age was 13.7% (37/270) in the avoidance group and 1.8% (5/272) in the consumption group.^[There were twelve further infants in this group whose results 
could not be evaluated.]
Among the 98 participants who initially had positive test results, the prevalence of peanut allergy was 35.3% (18/51) in the avoidance group 
and 10.6% (5/47) in the consumption group. There was no
between-group difference of consequence in the incidence of 
serious adverse events. 

In both groups, numbers and percentages are for those who were assigned 
to the group and whose results could be evaluated, whether or not they
followed the treatment protocol to which they were assigned.  In technical
terms, these are results from an "intention to treat" analysis. Such an
analysis is designed to mirror what can be expected in practice --- not
everyone who starts off in one group will stick to it.  It answers 
questions about what to do with subjects who did not fully follow
the treatment to which they were assigned.

The results were followed, in 2016, by changes to guidelines that 
recommended introduction of peanut and other allergenic foods 
before 12 months.  The assumption that avoiding early exposure to
peanuts would reduce risk of later development of peanut allergy
was, it was judged, likely wrong for all infants.

## False Positives

In contexts where the number of false positives is likely to be
high relative to the number of true positives, screening 
programs may have serious downsides that outweigh the benefits.
Smith [-@smith-sd, pp.98-99] gives the following example:

* A test for marijuana has a 95% accuracy (true positive) rate
* Of those who test positive, what fraction are marijuana users?

Administer test to 10,000 employees --- 500 use, 9500 do not

```{r marijuana, results="asis"}
tab <- matrix(c(475,475,950,25,9025,9050,500,9500,10000), ncol=3, nrow=3)
dimnames(tab) <- list(c("User","Not user","Total"),c("Test +ve","Test -ve","Total"))
tab <- xtable::xtable(tab, digits=0)
print(tab,type='latex', comment=FALSE, floating=FALSE, scalebox=1.0)
```

* Consider a test for excess iron syndrome that has ~80% accuracy  
Excess iron syndrome has the name "haemochromatosis".  Again,
administer it to 10,000 individuals, where 50 have the syndrome,
9950 do not (a rate of 1 in 200).

```{r iron, results="asis"}
tab <- matrix(c(40,1990,2030,10,7960,7970,50,9950,7000), ncol=3, nrow=3)
dimnames(tab) <- list(c("Haem...","Not haem...","Total"),c("Test +ve","Test -ve","Total"))
tab <- xtable::xtable(tab, digits=0)
print(tab,type='latex', comment=FALSE, floating=FALSE, scalebox=1.0)
```

### The Aspirin story (randomized trials) {-}

22,000 males, 1st 5 years of study:^[<http://nyti.ms/2rgjFO2>]

| |Placebo group | Aspirin group
|---|--------------|--------------
| Fatal heart attacks | 18 | 5
| Non-fatal heart attacks | 171 | 99

Later work: Benefits are limited to those with previous symptoms  
2009 study: Major bleeding episodes, requiring admission to hospital:  
were 20 (1.2%) in the placebo group; 34 (2%) in aspirin group^[<https://www.theguardian.com/society/2009/aug/31/aspirin-british-heart-foundation>]

For a summary of current evidence, see <http://mayocl.in/1cBM0ze>

## The effectiveness of surgery

The blurb on the back cover of @harris2016book states that

> For many complaints and condiltions, the benefits from surgery
are lower, and the risks higher, than you or your surgeon think.

Humans are very prone to the _post hoc, ergo propter hoc_ fallacy:
"it followed, therefore it was because of" fallacy.  Harris 
argues that unless the benefits of a surgical procedure are clear, 
the only ethical way forward is to do a randomized trial where 
the procedure is compared with a sham procedure.  Such trials
are not easy to design and execute. Nonetheless, there are a
number of important cases where such comparisons have been made.

Bloodletting is a prime example of a surgical procedure that 
has faded away due to evidence, not just of a lack of 
effectiveness, but of serious harm.^[For the history, see for 
example @seigworth1980bloodletting] The practice attracted 
widespread debate in the 19th century and into the early 20th 
century, with its defenders making such claims as

> "blood-letting is a remedy which, when judiciously employed, it is hardly possible to estimate too highly"


