# Matters of consequence

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment=NA, fig.align='center', fig.width=5.5,fig.height=5.5)
```

## [The MMR vaccine scandal](
https://en.wikipedia.org/wiki/MMR_vaccine_controversy)

The MMR vaccine was developed to for use in preventing measles, 
mumps, and rubella.  Andrew Wakefield was the lead author of a 
study published in 1998, based on just twelve children,  that 
claimed to find indications of a link between the MMR vaccine 
and autism. The journalist Brian Deer had a key role in 
identifying issues with the work, including fraudulent 
manipulation of the medical evidence.

It emerged that

* Wakefield had multiple undeclared conflicts of interest
* Funding came from a group of lawyers who were interested
    in possible personal injury lawsuits
* From 9 children said to have regressive autism   
    + Only 1 had been diagnosed; 3 had no autism
    + 5 had developmental problems before the vaccine

Wakefield's 1998 claims were widely reported

* Vaccination rates in the UK and Ireland dropped sharply
* The incidence of measles and mumps increased, resulting
  in deaths and in severe and permanent injuries.
    
Wakefield was found guilty by the General Medical Council of serious professional misconduct in May 2010 and was struck off the Medical Register.     
Following the initial claims in 1998, multiple large epidemiological studies failed to find any link between MMR and autism.

Fact boxes on the Harding site summarize evidence of the 
effectiveness of the [MMR vaccine](https://www.hardingcenter.de/de/search/node?keys=mmr)^[https://www.hardingcenter.de/de/search/node?keys=mmr]

## Sally Clark's disturbing cot death story 

Sudden Infant Death (SID), also referred to as "cot death",
is the name given to the unexplained sudden death of very
young children.  The story of Sally Clark's unfortunate brush 
with the law, following the death of a second child, is
interesting, disturbing, and educational.  Sally's experience 
highlighted ways in which the UK legal system needed to take 
on board issues that affect the use of medical evidence ---
issues of a type that can be important in medical research.

Paediatrician Sir Roy Meadow had argued in his 1997 book
**ABC of Child Abuse** that "unless proved otherwise, 
one cot death is tragic, two is suspicious and three is murder." 
Was this an example of "Too hard! Try something easier, and wrong!"
Or was it the triumph of assumed "knowledge" over hard evidence?

Meadow's role as an expert witness for the prosecution in several 
trials played a crucial part in wrongful convictions for murder.
The case that attracted greatest attention is that of Sally Clarke,
both of whose children were "cot death" victims.  Following the 
death of the second child, Meadow gave evidence at her 1999 trial, 
and appeal in 2000. Sally Clark was finally acquitted in 2003.

* Meadow gave 1 in 8,500 as cot death rate in affluent non-smoking families
    + Squared 8,500 to get odds of 73,000,000:1 against for two deaths.
    + Meadow assumed, wrongly, that the probability of a death from
    natural causes was the same in all families.
    + A first "cot death" is, in some families at least, evidence of a greater
    proneness to death from natural causes.
* Royal Statistical Society press release: "Figure has no statistical basis"
    + 2000 appeal judges: The figure was a
    "sideshow" that would not have influenced the jury's decision.
    + The appeal judges' statement was described by a leading QC, not
    involved in the case as: "a breathtakingly intellectually dishonest 
    judgment."
* 2003: It emerged that 2nd death was from bacterial infection  
    + In a second appeal, Sally Clark was freed.
* 2005: Meadow was struck from medical register
    + 2006: reinstated by appeal court --- misconduct fell short 
    of "serious"!

Meadow was in effect assuming, without evidence, the
absence of family-specific genetic or social factors that
make cot deaths more likely in some families than in others. 
Why was Meadow's assumption of independence not challenged 
in the 1999 trial? Issues of whether or not different pieces 
of evidence are independent are surely crucial to assessing
the total weight of evidence.  Meadow had only enough 
understanding of probability to be dangerous. 

Sally Clark was finally acquitted on her second appeal in
2003, with her sense of well-being damaged beyond repair.
The forensic evidence had been weak.  The web page
  \url{https://plus.maths.org/content/os/issue21/features/clark/index}
  has a helpful summary of the statistical issues.  It
  quotes a study that suggests that the probability of a
  second cot death in the same family is somewhere between 1
  in 60 and 1 in 130.  Even after this adjustment, the
probability of death from natural causes of two children in
the same family is low.  But so also is the probability that
an apparently caring mother from an affluent middle class
family, with no history of abuse, will murder two of her own
children. Those are the two probabilities that must be
compared.  Anyone who plans to work as a criminal lawyer
ought to understand this crucial point.  In a
large population, there will from time to time be two deaths
from natural causes in the one family.

A small number of appeals were subsequently launched against
other convictions where evidence of the same type had been
presented, most of them successful.  A further consequence
was that the law was changed such that no person could be 
convicted on the basis of expert testimony alone. 

 The @watkins2000conviction article **Conviction by mathematical 
 error?: Doctors and lawyers should get probability theory right** 
 provided a trenchant critique of the perils of allowing 
 non-statisticians to present unsound statistical arguments, with 
 no effective challenge.  
 
 > Guidelines for using probability theory in criminal cases are urgently needed. The basic principles are not difficult to understand, and judges could be trained to recognise and rule out the kind of misunderstanding that arose in this case.

They are perhaps more difficult than Watkins allowed.  Watkins
somewhat over-states his case.  Watkins was at the time 
Director of Public Health for Stockport. 

## The Reinhoff and Rogoff saga

Figure \@ref(fig:debt2gdp12) plots data, for 1946 -- 2009 
from 20 "advanced" countries, that underpinned the 2010 
paper "_Growth in Time of Debt_" by the two Harvard economic 
historians Reinhart and Rogoff [RR].  The paper [@reinhart2010growth]
has been widely quoted in support of economic austerity programs
internationally.  There was a huge stir, in the media and on the
blogosphere, when graduate student Herndon found and published 
details of coding and other errors in the results that RR had
presented.

```{r cap1and2, echo=FALSE}
cap3 <- "Dashed horizintal lines show means by Debt/GDP category, 
for 20 'advanced' countries, for the years 1946 --- 2009.
(Data are missing for some countries in some years.)
RR's mean (dots) for > 90\\% Debt/GDP was from 10 only of the 20. 
The smooth gray curve treats points as independent."
```

```{r GG-mgcv, echo=FALSE}
suppressPackageStartupMessages(library(ggplot2, quietly=TRUE))
suppressPackageStartupMessages(library(mgcv, quietly=TRUE))
```

```{r debt2gdp12, echo=FALSE, fig.width=7, fig.height=4.75, warning=FALSE, message=FALSE, out.width="100%", fig.align="center", fig.show="hold", fig.pos='H',fig.cap=cap3, cache=TRUE}
load('data/RR.RData')
RR.gam <- gam(dRGDP ~ s(debtgdp, bs="cs"),data=RR)
res <- residuals(RR.gam)
avTO30 <- mean(subset(RR, debtgdp<=30)[ ,'dRGDP'])
av30TO60 <- mean(subset(RR, debtgdp>30 & debtgdp<=60)[ ,'dRGDP'])
av60TO90 <- mean(subset(RR, debtgdp>60 & debtgdp<=90)[ ,'dRGDP'])
av4 <- sapply(split(RR,RR$Country), function(x)
  c(avTO30 = mean(subset(x, debtgdp<=90)[ ,'dRGDP']),
  av30TO60 = mean(subset(x, debtgdp>30 & debtgdp<=60)[ ,'dRGDP']),
  av60TO90 = mean(subset(x, debtgdp>60 & debtgdp<=90)[ ,'dRGDP']),
  av90plus = mean(subset(x, debtgdp>90)[ ,'dRGDP']),
  n90plus = nrow(subset(x, debtgdp>90))))
av4n <- av4[,!is.nan(av4[4,])]
av90plus <- sum(av4n[4,]*av4n[5,])/sum(av4[5,])
m <- ggplot(RR, aes(x=debtgdp,y=dRGDP), environment=environment())
m <- m + geom_vline(xintercept=90,color='lightgray',size=1.5)
m <- m + geom_point(color='darkgray', size=1.5) + ylab("Real GDP Growth") +
  xlab("Public Debt/GDP Ratio") +
  scale_x_continuous(breaks=seq(0,240,30)) + theme_bw()
m <- m + geom_segment(aes(x = 0, y = avTO30, xend = 30, yend = avTO30),
                      linetype=2, size=1, alpha=0.45, color="sienna") +
  geom_segment(aes(x = 30, y = av30TO60, xend = 60, yend = av30TO60),
               linetype=2, size=1, alpha=0.45, color="sienna") +
  geom_segment(aes(x = 60, y = av60TO90, xend = 90, yend = av60TO90),
               linetype=2, size=1, alpha=0.45, color="sienna") +
  geom_segment(aes(x = 90, y = 2.2, xend = 245, yend = 2.2),
               linetype=2, size=1, alpha=0.45, color="sienna") +
  geom_segment(aes(x = 90, y = -0.1, xend = 245, yend = -0.1),
               linetype="dotted", size=1, alpha=0.45, color="black")
keep <- (res > 11.15)|(res < -10.15) | (RR[,'debtgdp'] > 210) |
  (RR[,'debtgdp'] > 90 & abs(res)>5) | (RR[,'debtgdp'] > 180 & res<0)
RRkeep <- RR[keep,]
hj <- rep(-0.085, sum(keep))
onleft <- with(RRkeep,(debtgdp < 90)|
                 (debtgdp > 220))
hj[onleft] <- 1.075
txlab <- with(RRkeep, paste(iso2c, Year))
m <- m + geom_smooth(method=gam, color=adjustcolor('black',alpha=0.45),
                     formula= y ~ s(x, bs = "cs"),lwd=1.5)
m1 <- m + geom_text(data = RRkeep, aes(x=debtgdp,y=dRGDP, label = txlab),
                    size=3, hjust = hj)+xlab("Ratio of debt to GDP (%)")+
                    ylab("GDP growth (%)") +
                    ggtitle(paste("GDP growth vs Debt/GDP ratio",
"-- For >90% ratio, cf dashed vs dotted lines"))
tagged <- as.character(unique(RRkeep$Country))
tag2 <- RRkeep[match(tagged, RRkeep[, 'Country']), 'iso2c']
tag2 <- as.character(tag2)
leg <- cbind(tag2, tagged)[tag2!=tagged,]
dimnames(leg)[[2]] <- NULL
library(gridExtra)
tt <- ttheme_default(base_size=9)
print(m1+annotation_custom(tableGrob(leg, theme=tt),
                           xmin=210, xmax=250,
                           ymin=10.0, ymax=28))
```

As well as coding errors, 
[Herndon](http://www.peri.umass.edu/fileadmin/pdf/working_papers/working_papers_301-350/WP322.pdf)^[<http://www.peri.umass.edu/fileadmin/pdf/working_papers/working_papers_301-350/WP322.pdf>] identified selective exclusion of available data, 
and unconventional weighting of summary statistics.  There was a 
failure to acknowledge that the relationship studied has varied 
substantially by country and over time.   Half of the 20 countries
had missing data for 5 or more of the years, with the largest
number missing in the years 1946 to 1949.

In response to @herndon2014does and other critics, Reinhart and 
Rogoff accepted that coding errors had led to the omission of 
several countries, but pushed back against other criticisms. 
Their revised analysis addresses only the most egregious errors 
in their work.  Among other issues, their insistence on treating 
each data point for each country as an independent piece of 
evidence makes no sense.  The smooth curve fitted in Figure
\@ref(fig:debt2gdp12) may be regarded as an average over the
ten countries, but (as the next graph will show)  with huge country
to country variation.

```{r cap3, echo=FALSE}
cap3 <- "Smooths have been fitted for each of the 10 countries
for which debt to GDP ratios were in some years greater than 90\\%.
There is no consistent pattern, as there should be if RR's claim
is to hold up."
```

```{r debt2gdp-3, echo=FALSE, fig.width=7.0, fig.height=3.5, warning=FALSE, error=FALSE, fig.pos="H", fig.cap=cap3}
  if(!require(lattice, quietly=TRUE))stop("Lattice package must be available.")
trellis.par.set(theme)
RR10 <- droplevels(subset(RR, Country%in%colnames(av4n)))
fit.gam <- gam(dRGDP~s(debtgdp, bs='cr', k=4, by=Country), 
               family=scat, data=RR10)
hat <- fitted(fit.gam)
gph4 <- xyplot(hat~debtgdp, groups=Country, data=RR10, type="l", 
                ylim=c(-0.5,7.5),
 xlab="Ratio of debt to GDP (%)", ylab="GDP growth (%)", out.width='80%',
 panel=function(x,y,groups,subscripts,...){
 panel.superpose(x, y, ...,
 panel.groups=function(x,y,...){
 ord <- order(x)
 xord <- x[ord]
 yord <- y[ord]
 panel.lines(xord,yord, ...)
 im <- which.max(xord)
 xm <- min(xord[im],240)
 ym <- yord[im]
 p <- rep(4, length(ym))
 gp <- list(...)$group.value
 if(gp%in%c("Canada"))ym <- ym+0.25
 if(gp%in%c("Belgium"))ym <- ym-0.1
 if(gp%in%c("Ireland"))ym <- ym-0.05
 if(gp%in%c("US"))ym <- ym-0.2
 if(gp%in%c("Italy"))ym <- ym-0.25
 if(gp%in%c("Japan"))ym <- ym+0.025
 if(gp%in%c("Australia"))ym <- ym-0.05
 if(gp%in%"Ireland") p <- 2
 panel.text(xm,ym,labels=gp, pos=p, offset=0.25)
 }, groups=groups, subscripts=subscripts)
   })
## gph3 <- xyplot(dRGDP ~ debtgdp, groups=Country, data=RR10)
gph4
```

### Is there a pattern across countries? {-}

```{r av4}
av4 <- sapply(split(RR,RR$Country), function(x)
c(avTO30 = mean(subset(x, debtgdp<=90)[ ,'dRGDP']),
av30TO60 = mean(subset(x, debtgdp>30 & debtgdp<=60)[ ,'dRGDP']),
av60TO90 = mean(subset(x, debtgdp>60 & debtgdp<=90)[ ,'dRGDP']),
av90plus = mean(subset(x, debtgdp>90)[ ,'dRGDP']),
nyears = nrow(subset(x, debtgdp>90))))
av4n <- av4[,!is.nan(av4[4,])]
dRGDPav <- sum(av4n[4,]*av4n[5,])/sum(av4[5,])
```
There were just `r ncol(av4n)` countries where debt to GDP ratios were 
greater than 90% for one or more years.  For `r sum(av4n[4,]>0)` of 
those countries, the average of their GDP growth in the years at issue 
was on average positive relative to the previous year.  The average 
percentage growth over all `r ncol(av4n)` countries, weighted according 
to number of years, was `r round(dRGDPav, 3)`.

### There are further serious issues of interpretation {-}

* Does GDP drive debt/GDP ratio, or is it the other way round?
    + Or does a third factor drive both?
* Is the effect immediate, or on future economic performance

Smith (p.64) refers to work indicating that economic performance
is more closely correlated with economic growth in the past than
with future growth.

### Parting comments {-}

@herndon2014does comment

"... RR’s findings have served as an intellectual bulwark in support of austerity politics. The fact that RR’s findings are wrong should therefore lead us to reassess the austerity agenda itself in both Europe and the United States."  
  
The saga emphasizes the importance of working with reproducible code, 
rather than with spreadsheet calculations. 
The errors in RR's calculations were from one perspective fortunate.
Once highlighted, the errors drew critical attention to the paper, 
and to the serious flaws in the analysis.

<!-- ## The "pills may cause blood clots" saga -->

<!-- An October 1995 letter to doctors from the UK Committee -->
<!-- on the Safety of Medicines cited "new evidence" -->
<!-- indicating that:   -->
<!-- > . . . __the risk of thrombosis is increased by a factor of -->
<!-- two for some contraceptives relative to others__ . . .  -->
<!-- > but the pill is safe for most women . . . [Ellenberg, p.118] -->
<!-- Associated Press Report: http://bit.ly/2pQOxXQ -->

<!-- The change in risk was small.  The estimated risk increase was from 1 in 7000 -->
<!-- to 2 in 7000 -->

<!-- * 2 $\times$ 0.000014 = ?  (Risk of death was smaller still)     -->

<!-- A later analyses gave the risk as 1.7 in 7000, i.e., no increase, -->
<!-- irrespective of the contraceptive used.  There were other  -->
<!-- consequences that had not been adequately considered. -->
<!-- Comparing 1996 with 1995, there were 26,000 more births, and 13,600 -->
<!-- more abortions.  A large part of this increase was surely a result -->
<!-- of the "Dear Doctor" letter. -->

## What do malaria drugs do to Covid-19 patients?^[Lancet, May 2020, https://bit.ly/3xqncMt] {#sec:lancet}

Thirteen days after it was published on May 20 2020, three of the 
four authors withdrew a paper that claimed to find that malaria drugs, 
when used experimentally with patients with Covid-19, led to around 
30% excess deaths.  Irrespective of the problems with the data that
will be noted shortly, serious flaws in the analysis ought to have
attracted the attention of referees.  There was inadequate 
adjustment for known and measured confounders (disease severity, 
temporal effects, site effects, dose used).

The study claimed to be based on data from 96,032 
hospitalized COVID-19 patients from six continents, of which 66%
were from North Ammerica.  Very soon after it appeared, the article 
attracted critical attention, with a number of critics joining 
together to submit the letter @watson2020open to Lancet.

The sources from which the data had been obtained could not be verified,
data that claimed to be from just five Australian sources had more
cases than the total of Australian government figures, and similarly
for Australian deaths, there were implausibly small reported variances
in baseline variables, mean daily doses of hydroxychloroquine that were 
100 mg higher than US FDA recommendations.

Randomized trials designed to test the effectiveness of the drugs,
and that were in progress at the time when the paper appeared,
were temporarily halted.  The eventual conclusion was that the
drugs did not improve medical outcomes. There was some evidence
that hydroxychloroquine could have adverse effects.

With current web-based technology, RCTs can be planned and carried out and yield definitive answers, in much the same time as it would take to collect and analyze the data that are required for an observational study whose conclusions can be, at best, suggestive. Data confidentiality issues are easier to handle in the context of an RCT.

## A simplistic use of publicly available data
A June 2021 paper in _Vaccines_, titled 
"The Safety of COVID-19 Vaccinations -— We Should Rethink the Policy."^[_Vaccines_, June2021, bit.ly/3dTg1oh]
massively over-stated vaccine risk.  The paper claimed that
“For three deaths prevented by vaccination we have to accept 
two inflicted by vaccination.” 

* Deaths were from any cause, post-vaccination, reported both by 
professionals and the public  
    (Such data has to be used with a 'baseline' comparison)
* Vaccine benefits extend far beyond 6 weeks of Israeli study
* Focused on immediate risk to individual, not community.
* The way that any risk from the vaccine balances out against
risk of death from Covid-19 will vary depending on the the age 
structure of the population, on proportion immunized within 
each age group, and on the age and health of the individual. 

Deaths that could be verified as from the Pfizer vaccine have been, 
with the possible exception of frail elderly people, extremely rare.
See US CDC report on risk (for Pfizer, anaphylaxis)^[www.cdc.gov/coronavirus/2019-ncov/vaccines/safety/adverse-events.html],
and Helen Petousis-Harris' commentary on risks.^[sciblogs.co.nz/diplomaticimmunity/2021/04/15/covid-vaccines-and-blood-clots-what-is-this-about/]


