---
output:
  pdf_document: default
  html_document: default
---
# Notes

## 1. The @jung2014female US hurricane data {#hurricanes .unnumbered}

Figures \@ref(fig:hurricanes1) and \@ref(fig:hurricanes2) checked 
for a difference between the fitted male and female line or curve.
Jung et al's approach was, instead, to examine whether numbers of 
deaths varied with the "femaleness" of name, as judged by students 
in 2014.  

As a check on how the popularity of a name for each of females 
and males may have changed with time, Table \@ref(tab:changetab)
uses US social security administration data to show numbers for
names where the range of variation over 1950 to 2012 in the 
relative number of females was 0.08 or more.

```{r cap1, echo=FALSE}
cap1 <- "The minimum and maximum value of the relative proportion
of female names, and the difference, are shown for the eight names
that showed the greatest change over the years 1950 to 2012."
```

```{r changetab, out.width="90%", fig.align='center', echo=FALSE}
mat <- match(babynames::babynames$name, DAAG::hurricNamed$Name, nomatch=0)
matnames <- babynames::babynames[mat>0,]
mf <-merge(subset(matnames,sex=='F')[, c(1,3:4)], 
           subset(matnames,sex=='M')[,c(1,3:4)],all=TRUE,
           by=c("year",'name'),suffixes=c('f','m'))
mf <- within(mf, {nf[is.na(nf)]<-0; nm[is.na(nm)]<-0})
mf$fracf <- with(mf, nf/(nf+nm))
## table(mf$name[mf$fracf>0.075&mf$fracf<0.925])
mfsub <- subset(mf, year%in%(1950:2012))
ran <- with(mfsub, sapply(split(fracf,name),range))
d <- apply(ran, 2, diff)
ord <- round(order(abs(d), decreasing=T))
stats <- round(rbind(ran[,ord[1:8]],d[ord[1:8]]),2)
rownames(stats) <- c('Minimum','Maximum','Difference')
z <- match(DAAG::hurricNamed$Name, mfsub$name, nomatch=0)
library(kableExtra)
kable(stats, digits=2, caption=cap1) %>% 
  row_spec(3, bold=TRUE) %>% kable_styling(font_size=10)
```
In other cases, relative number of females were always
either in the range 0 to 0.07 (i.e., mostly female), or
in the range 0.93 to 1 (i.e., mostly female).

Figure \@ref(fig:plotchanges) shows how the numbers of
each sex changed over time, for the first six of the
names in Table \@ref(tab:changetab).

```{r cap2, echo=FALSE}
suppressPackageStartupMessages(library(latticeExtra))
trellis.par.set(theme)
cap2 <- "Change in numbers of names given to males and females
over the years 1950 to 2012, for the six names where the 
maximum difference in relative frequency was more than 0.1.
The maximum change is shown against each name."
```


```{r plotchanges, echo=FALSE, eval=TRUE, fig.width=5.5, fig.height=3.6, out.width="90%", fig.cap=cap2, fig.align='center'}
lookat <- subset(mfsub, name%in%c('Charley','Cleo',
                               'Fran','Sandy','Erin','Inez'))
lookat$name <- factor(lookat$name)
dnam <- round(abs(d[c('Charley','Cleo','Fran','Sandy','Erin','Inez')]),2)
levels(lookat$name) <- paste0(levels(lookat$name), ": ",dnam)
atx <- c(10,100,1000, 10000)
xyplot(nm+nf~year|name, data=lookat,  xlab="Year", 
       ylab="Number of babies", par.settings=simpleTheme(pch=c(3,2), cex=0.8), 
       auto.key=list(text=c("Males","Females"),columns=2),
       scales=list(y=list(log='e', at=atx, labels=paste(atx)))) +
  latticeExtra::layer(panel.axis('left', at=atx*sqrt(10), labels=F, tck=0.5))
```

The names where the choices of parents at the time are likely to
most different from that for students in 2014 are those where
there has been greatest change over time (i.e., especially, 
Charley, Cleo and Fran).

\enlargethispage{21pt}

Other differences from the analyses on which Figures 
\@ref(fig:hurricanes1) and \@ref(fig:hurricanes2) were based
are

* As the primary measure of the risk posed by the hurricanes, the 
authors used a 2013 US\$ estimate of damage, intended for
insurance purposes, for a comparable hurricane in 2013.  Figure 
\@ref(fig:hurricanes1) uses what is surely the more relevant
measure, namely `NDAM2014` --- this converts the estimate of
damage caused at the time to 2014 US\$.^[Fortuitously, this 
change makes no difference of consequence to the graph or to 
the conclusions reached.] 
* Jung et al allowed for minor effects from barometric pressure 
at landfall, and interactions. Again, these do not affect 
the conclusions reached.

```{r missed, echo=FALSE, eval=FALSE}
missed <- DAAG::hurricNamed$Name[z==0]
mfOther <- subset(mf, name%in%missed)
ranOther <- with(mfOther, sapply(split(fracf,name),range))
dOther <- apply(ranOther, 2, diff)
```

```{r changes, echo=FALSE}
stats <- round(rbind(ran[,ord[1:8]],d[ord[1:8]]),2)
x <- knitr::kable(stats)
```

## 2. $^*$What does a $p$-value tell the experimenter? {#pval .unnumbered}

What follows introduces technicalities that have been avoided in the
preceding chapters. 

Consider an experiment that compares a treatment with a control.
A $p$-value measures the probability that a value as small as
that observed, or smaller, would occur by chance if there really
was no difference in effect between treatment and control.
(The assumption of no difference is termed the NULL hypothesis.)
The probability of 0.05, or 1 in 20 applies in the case where
a threshold of 0.05 has been set in advance --- all values
under that threshold are lumped together without distinction.

Issues for understanding what a $p$-values tells you are:

* It does not tell you what happens if there is a difference.
    + In particular, it does not tell you the probability that 
    there really is a difference.
* Finding $p \leq 0.05$ is all very well.  What one really 
needs to know is that there is an alternative that is
substantially more likely.
* $p = 0.05$, or close to 0.05, is at the upper end of the range
of $p$-values that occur with a probability of 0.05.  Treating
$p = 0.05$ as an event that occurs with probability 0.05 under
the NULL greatly exaggerates the evidence that it provides 
against the NULL.
    + In fact, under the NULL, all values between 0 and 1 are
    equally likely!

