--- 
title: 'What Do the Data Say? -- Traps  to Avoid'
subtitle: "Examples that inform and educate"
author: "John Maindonald"
date: "11 April 2024"
site: bookdown::bookdown_site
documentclass: book
classoption: b5paper
fontsize: 10ptls
urlcolor: blue
linkcolor: red
bibliography: [book.bib, packages.bib, dataSense.bib]
biblio-style: apalike
link-citations: true
github-repo: jhmaindonald/dataThoughts
description: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment=NA, fig.align='center', fig.width=5.5,fig.height=5.5)
```

\renewcommand{\bibname}{References}

# Preface {-}

> It ain't what you don't know that gets you into trouble.
> Its what you know for sure that just ain't so.
> [Variously attributed; author unknown]

This booklet has as its intended audience, as well as practicing 
researchers, anyone interested in using data as a basis for
judgment.  The critical processes and skills that are discussed 
have wide application, in everyday as well as in professional life.

The questions that data and data analysis may be asked to answer can 
often be stated simply. This may encourage the layperson to believe 
that the steps needed to provide answers are similarly simple.
Very often, they are not.  It is alluringly easy to create forms of
data summary that misrepresent what the data have to say.  
Or inadequate attention may be paid to the context from which the
data have been taken.  The context has to be understood and to feed 
into the way that the data are used, if conclusions are to be drawn
that warrant credence.

The pages that follow avoid detailed discussion of methodology,
instead focusing on what Kahneman, in his book 
"Thinking Fast and Slow"^[@kahneman_2013] calls "educating gossip", 
with examples taken from the media and from the research literature.
Examples are chosen for the insight that they may provide, both
those that show effective critical processes at work, and those
that do not.

When they function well, scientific processes work to avoid the
traps that will be discussed.  They fail often enough that examples 
of failure are relatively easy to find.  Cases where scientific 
processes have clearly failed are commonest where scientists work 
as individuals or in small groups with limited outside checks.

In areas where the nature of the work requires cooperation between 
scientists with a wide range of skills, and where data and code
are shared, those involved in the research provide a check on each 
other.  Papers may be sent for comment to other researchers, or 
posted on the net for comment, prior to formal submission for 
publication.  This allows informed and incisive criticism, with 
the more formal refereeing process providing supplementary checks.

Concerns about reproducibility, especially in wet laboratory 
biology and in psychology, have in the past decade attracted 
wide attention in the pages of Nature, Science, the Economist, 
psychology journals, and elsewhere.  Among other needed changes, 
publication of experimental results needs to become a staged 
process that moves from "this is worth a look" to "has been 
independently replicated" to "established result".

Human psychology helps explain why humans so readily fall for
conclusions that are both simplistic and wrong. Humans are 
programmed, by inheritance and by conditioning, to respond 
quickly to signs of immediate danger.  Too often, we respond 
quickly and without careful thought in situations that call 
for a carefully reasoned response.  Or, we may not have the 
skills needed for a carefully reasoned judgment.
It is helpful to think about the psychology involved when humans 
make judgments as of two types, which Kahneman (2013) calls System 1 
(jumping quickly to a conclusion) and System 2 (pondering). 
These are a useful starting point for drawing attention to common 
traps to which humans are prone. Those who know and understand 
common traps are better placed to avoid them.

# Summary of points that are discussed  {-}

1. Understand the psychology  --- why the showcase of fallacies
    +  Recognize and avoid the mental traps to which humans are prone.
2. Graphs can reveal surprises. They can also be drawn 
so that they deceive.  
    + Choice of one or both scales such that equal relative
    changes mark out equal distances on the scale may show 
    patterns that are not otherwise obvious.
3. From hindsight to insight --- it is easy to be deceived.
    + If a sportsperson is at the top of their form, the only 
way to go is down. If at the bottom, the only way is up.
    + This is true also for success in business.
    + Chance, as well as form, obviously plays a part.
4. Medicine and health provides a good context into which to study
a number of important issues. What causes cholera – bad air or bad water? 
 Does drinking coffee help or harm health? 
    + Randomized trials, if done rigorously, and participants 
    closely reflect the population to which results will be applied,
    are a gold standard.
    + Think carefully about the outcome measure.  Thus, is it
    "cancers" found?  Or is it "deaths" within a stated timeframe.
5. Careful checks are needed when the attempt is made to use
observational data to establish causation.
    + In group comparisons, adjustments are needed 
    to account for prior differences between the two (or more)
    groups. What checks can be made that adjustments are adequate?
    It is in general impossible to be sure that adjustments account
    for all effects other than the effectm that is of interest.
    + In what direction, and through what causal chains, do causal
    effects go? Are people healthier because they exercise more?
    Or, do they exercise more because they are healthier?  Or, do
    causal effects go in both directions?  What are the influences
    from dietary and other lifestyle factors?
    + An overwhelming case can sometimes be made, as in the link
    between smoking and lung cancer, by bringing together multiple 
    independent lines of evidence.
6. Mistakes that arise from statistical misunderstandings
can have serious consequences.
    + This is the case also for fraudulent manipulation of
    data and/or evidence.
7. Weighting paradoxes appears in many different guises.
A notable example is the Yule-Simpson paradox, which arises 
because we have naively added numbers in ways that give more 
weight to some combinations of effects than to others. 
8. What direction does the correlation go? 
    + Tall fathers are likely to have tall sons, but shorter than themselves. 
    + What can be hard to grasp is that the effect goes on both directions.
Tall sons are likely to have tall fathers, but shorter than themselves.
The "regression to the mean" phenomenon goes in both directions.  
    + Regression models require careful checking to ensure that they
    correctly reflect all systematic effects in the data.
    + The coefficient that is attached to an explanatory variable accounts
    for the effect of that variable when other variables are help constant.
    The coefficient is likely to change with the change to different 
    explanatory variables that give the same predicted  values.
9. Results become part of established science when they have
survived informed critique. Work that claims to revise or overturn
established results has itself to survive the same informed critique.
    + Uninformed critique is common.  Watch for the tricks that 
    detractors who find established scientific results inconvenient
    use in the attempt to discredit them.
    + When presented with scientific (or any) claims, ask/check
    whether claims have been carefully critiqued.  For laboratory
    studies, have results been replicated?
    + Publication processes work best where authors contribute
    input and critique widely from across different disciplinary 
    perspectives, and where data and code are made available for 
    checking.


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'MASS', 'mgcv', 
  'GDAdata','corrplot','latticeExtra','DAAG','xtable'
), 'packages.bib')
if(!file.exists("book-index.ind"))
  system("cp '_book/work/book-index.ind' 'book-index.ind'")
```
